Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-05-13 content.tex
Improve Table 3, use 'Table' and 'Figure' when referring to tables and figures

\verb;pirouette; is very flexible and allows the user 
to specify a wide variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in tables~\ref{tab:options} and~\ref{tab:definitions}.

@@

  \begin{tabular}{|@{}c|p{4cm}|p{12cm}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} &
    \textbf{Macro-argument} &
    \textbf{Description} \\
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{G}$ &
    Generative model &
    The full setting to produce BEAST2 input data. 
    Its core features are the tree prior $\mathit{p_{G}}$, the clock 
    model $\mathit{c_{G}}$ and the site model $\mathit{s_{G}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  
    $\mathit{A}$ &
    Alignment model &
    It includes the parts of the generative model that directly affect the 
    alignment generation, like the clock model $\mathit{c_{G}}$ and 
    the site model $\mathit{s_{G}}$. Additional arguments can be provided, 
    such as the mutation rate and the root sequence. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{X_{i}}$ &
    $i$-th candidate experiment &
    Full setting for a Bayesian inference. It is made by a 
    candidate inference model $\mathit{I_{i}}$ and its 
    inference conditions $\mathit{C_{i}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{I}$ &
    Inference model &
    Phylogenetic inference model to run BEAST2. Likewise the generative model $G$, 
    its main components are the tree prior $\mathit{p_{I}}$, the 
    clock model $\mathit{c_{I}}$ and the site model $\mathit{s_{I}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ 
    is used in the inference. 
    They are composed by the model type, run condition and 
    wether to measure the evidence. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{E}$ & Error measure parameters & 
    Errors measurement setup that can be specified providing an 
    error function to measure the difference between the original phylogeny 
    and the inferred posterior. The initial part of the posterior that is 
    reckoned as not representative can be discarded using a burn-in fraction. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \iffalse
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure & 
      $\mathit{\tau_{p}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \fi
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in the main text and in 
    Fig~\ref{fig:pipeline}. To run the pipeline $\mathit{A}$, $\mathit{X}$ 
    and $\mathit{E}$ must be specified. Examples can be found in 
    listings~\ref{lst:create_alignment}, 
    \ref{lst:create_gen_experiment_explicit} and 
    \ref{lst:create_error_measure_params}.
  }
  \label{tab:definitions}
\end{sidewaystable}

@@

An overview of \verb;pirouette;'s main functions is shown in 
table~\ref{tab:functions}. 
Their usage is demonstrated in the example code below.
All \verb;pirouette;'s functions are documented,
have a useful example and sensible defaults.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
  \centering
  \begin{tabular}{ | l | l | }
    \hline
    \textbf{Name} & \textbf{Description} \\
    \hline
    \verb;pir_run; & Run \verb;pirouette; \\
    \verb;pir_plot; & Show the \verb;pirouette; results as a plot \\
    \verb;create_pir_params; & Create the \verb;pirouette; parameters \\
    \hline
    \verb;create_alignment_params; & Create the alignment parameters \\
    \verb;create_twinning_params; & Create the twinning parameters \\
    \verb;create_experiment; & Create one experiment \\
    \verb;create_error_measure_params; & Create the error measurement parameters \\
    \hline
  \end{tabular}
  \caption{pirouette's main functions.}
  \label{tab:functions}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{lstlisting}[
    language = R,
    floatplacement = ht,
    frame = single, 
    label = {lst:create_yule_tree}, 
    caption = {
      Create a Yule tree. 
      The resulting tree is shown in figure~\ref{fig:yule_tree}.
    }
  ]
phylogeny <- create_yule_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}

@@

\begin{lstlisting}[
    language = R,
    floatplacement = ht,
    frame = single,
    label = {lst:create_alignment}, 
    caption = {Create an alignment.}
  ]
alignment_params <- create_alignment_params(
  clock_model = create_strict_clock_model(),
  site_model = create_jc69_site_model(),
  mutation_rate = 0.1,
  root_sequence = create_blocked_dna(length = 1000)
)
\end{lstlisting}

@@

The resulting error distribution, as shown figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it from ERQ 3.

@@

The first step is to simulate a DNA alignment from the 
given phylogeny. 
We will re-use the alignment parameters of the previous example 
as shown in listing~\ref{lst:create_alignment}.

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
as shown in figure~\ref{fig:example_2}.

@@

Most of the settings for this ERQ are
the same as in ERQ 2: we will re-use the phylogeny from an 
unknown speciation model (listing~\ref{lst:unknown_phylogeny}), 
the alignment parameters (listing~\ref{lst:create_alignment}).

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
The output is shown in figure~\ref{fig:example_3}

@@

Here, as in ERQs 2 and 3, we start with a phylogeny 
generated by an unknown speciation model (listing~\ref{lst:unknown_phylogeny}). 
The first step is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

We specify that we want all 40 inference models to compete 
by calling \verb;create_all_experiments;,
as we previously did in listing~\ref{lst:create_all_experiments}.
We choose again to use the default error measurement setup,
as shown in listing~\ref{lst:create_error_measure_params}.
To get an idea of a baseline error made by BEAST2, 
we also make use of the twinning option by using
the same call to \verb;create_pir_params; as shown in 
listing~\ref{lst:create_pir_params_with_twinning}.
We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in figure~\ref{fig:example_4}.

@@

In this example we use a tree generated by the Yule process.
Instead of using the common Yule tree as in ERQ 1, we will use
a Yule tree of low likelihood, 
i.e. the one already used in ERQs 2-4, 
by using the code shown in listing~\ref{lst:unknown_phylogeny}.
The first step is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in figure~\ref{fig:example_5}.

@@

However, if the same procedure were repeated and performed on a distribution 
with a sufficient number of generative trees, 
it could actually constitute a quantitative assessment 
of the quality of the inference.
To give an idea of the insights \verb;pirouette; can give,
we pretend the results of the example research questions
to be obtained from thousands of replicates.
In that case, figure~\ref{fig:example_1} would show 
the minimum error BEAST2 when the
generative model is known, which can serve as 
a minimum noise level assessment in a bigger experiment.  
Figure~\ref{fig:example_2} shows that 
the best candidate model gives an error distribution similar to the unknown 
tree prior,
hinting that the candidate model would be just as good in inference.
Figure~\ref{fig:example_3} shows that 
the twin model gives an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2.
Figure~\ref{fig:example_4} shows that 
the twin model and best candidate model give 
an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2, 
and that the candidate model would be just as good in inference.

>>>>

\verb;pirouette; is very flexible and allows the user 
to specify a wide variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in Table~\ref{tab:options} and Table~\ref{tab:definitions}.

@@

  \begin{tabular}{|@{}c|p{4cm}|p{12cm}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} &
    \textbf{Macro-argument} &
    \textbf{Description} \\
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{G}$ &
    Generative model &
    The full setting to produce BEAST2 input data. 
    Its core features are the tree prior $\mathit{p_{G}}$, the clock 
    model $\mathit{c_{G}}$ and the site model $\mathit{s_{G}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  
    $\mathit{A}$ &
    Alignment model &
    It includes the parts of the generative model that directly affect the 
    alignment generation, like the clock model $\mathit{c_{G}}$ and 
    the site model $\mathit{s_{G}}$. Additional arguments can be provided, 
    such as the mutation rate and the root sequence. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{X_{i}}$ &
    $i$-th candidate experiment &
    Full setting for a Bayesian inference. It is made by a 
    candidate inference model $\mathit{I_{i}}$ and its 
    inference conditions $\mathit{C_{i}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{I}$ &
    Inference model &
    Phylogenetic inference model to run BEAST2. Likewise the generative model $G$, 
    its main components are the tree prior $\mathit{p_{I}}$, the 
    clock model $\mathit{c_{I}}$ and the site model $\mathit{s_{I}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ 
    is used in the inference. 
    They are composed by the model type, run condition and 
    wether to measure the evidence. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{E}$ & Error measure parameters & 
    Errors measurement setup that can be specified providing an 
    error function to measure the difference between the original phylogeny 
    and the inferred posterior. The initial part of the posterior that is 
    reckoned as not representative can be discarded using a burn-in fraction. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \iffalse
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure & 
      $\mathit{\tau_{p}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \fi
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in the main text and in 
    Fig~\ref{fig:pipeline}. To run the pipeline $\mathit{A}$, $\mathit{X}$ 
    and $\mathit{E}$ must be specified. Examples can be found in 
    listings~\ref{lst:create_alignment_params}, 
    \ref{lst:create_gen_experiment_explicit} and 
    \ref{lst:create_error_measure_params}.
  }
  \label{tab:definitions}
\end{sidewaystable}

@@

An overview of \verb;pirouette;'s main functions is shown in 
Table~\ref{tab:functions}. 
Their usage is demonstrated in the example code below.
All \verb;pirouette;'s functions are documented,
have a useful example and sensible defaults.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
  \centering
  \begin{tabular}{ | l | l | l | }
    \hline
    \textbf{Name} & \textbf{Description} & \textbf{Listing} \\
    \hline
    \verb;pir_run; & Run \verb;pirouette; & \ref{lst:pir_run} \\
    \verb;pir_plot; & Show the \verb;pirouette; results as a plot & \ref{lst:pir_plot} \\
    \verb;create_pir_params; & Create the \verb;pirouette; parameters & \ref{lst:create_pir_params} \\
    \hline
    \verb;create_alignment_params; & Create the alignment parameters & \ref{lst:create_alignment_params} \\
    \verb;create_twinning_params; & Create the twinning parameters & \ref{lst:create_twinning_params} \\
    \verb;create_experiment; & Create one experiment & \ref{lst:create_gen_experiment_explicit} \\
    \verb;create_error_measure_params; & Create the error measurement parameters & \ref{lst:create_error_measure_params} \\
    \hline
  \end{tabular}
  \caption{
    \texttt{pirouette}'s main functions, description and the number of the 
    listing in which it is used. 
  }
  \label{tab:functions}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{lstlisting}[
    language = R,
    floatplacement = ht,
    frame = single, 
    label = {lst:create_yule_tree}, 
    caption = {
      Create a Yule tree. 
      The resulting tree is shown in Figure~\ref{fig:yule_tree}.
    }
  ]
phylogeny <- create_yule_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}

@@

\begin{lstlisting}[
    language = R,
    floatplacement = ht,
    frame = single,
    label = {lst:create_alignment_params}, 
    caption = {Create an alignment.}
  ]
alignment_params <- create_alignment_params(
  clock_model = create_strict_clock_model(),
  site_model = create_jc69_site_model(),
  mutation_rate = 0.1,
  root_sequence = create_blocked_dna(length = 1000)
)
\end{lstlisting}

@@

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it from ERQ 3.

@@

The first step is to simulate a DNA alignment from the 
given phylogeny. 
We will re-use the alignment parameters of the previous example 
as shown in listing~\ref{lst:create_alignment_params}.

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
as shown in Figure~\ref{fig:example_2}.

@@

Most of the settings for this ERQ are
the same as in ERQ 2: we will re-use the phylogeny from an 
unknown speciation model (listing~\ref{lst:unknown_phylogeny}), 
the alignment parameters (listing~\ref{lst:create_alignment_params}).

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
The output is shown in Figure~\ref{fig:example_3}

@@

Here, as in ERQs 2 and 3, we start with a phylogeny 
generated by an unknown speciation model (listing~\ref{lst:unknown_phylogeny}). 
The first step is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment_params}.

We specify that we want all 40 inference models to compete 
by calling \verb;create_all_experiments;,
as we previously did in listing~\ref{lst:create_all_experiments}.
We choose again to use the default error measurement setup,
as shown in listing~\ref{lst:create_error_measure_params}.
To get an idea of a baseline error made by BEAST2, 
we also make use of the twinning option by using
the same call to \verb;create_pir_params; as shown in 
listing~\ref{lst:create_pir_params_with_twinning}.
We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_4}.

@@

In this example we use a tree generated by the Yule process.
Instead of using the common Yule tree as in ERQ 1, we will use
a Yule tree of low likelihood, 
i.e. the one already used in ERQs 2-4, 
by using the code shown in listing~\ref{lst:unknown_phylogeny}.
The first step is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment_params}.

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_5}.

@@

However, if the same procedure were repeated and performed on a distribution 
with a sufficient number of generative trees, 
it could actually constitute a quantitative assessment 
of the quality of the inference.
To give an idea of the insights \verb;pirouette; can give,
we pretend the results of the example research questions
to be obtained from thousands of replicates.
In that case, Figure~\ref{fig:example_1} would show 
the minimum error BEAST2 when the
generative model is known, which can serve as 
a minimum noise level assessment in a bigger experiment.  
Figure~\ref{fig:example_2} shows that 
the best candidate model gives an error distribution similar to the unknown 
tree prior,
hinting that the candidate model would be just as good in inference.
Figure~\ref{fig:example_3} shows that 
the twin model gives an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2.
Figure~\ref{fig:example_4} shows that 
the twin model and best candidate model give 
an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2, 
and that the candidate model would be just as good in inference.
